# Cross-references {#cross}

Cross-references make it easier for your readers to find and link to elements in your book. In Bookdown, you can create cross-references for chapters, sub-chapters, figures, and tables.

## Chapters and Sub-Chapters

There are two steps to cross-reference any heading:

1. **Label the heading**:  
   Add a label by placing `{#your-label}` at the end of the heading:
   ```markdown
   # Hello World {#nice-label}
   ```
   - If you don’t add a label, Bookdown will automatically generate one based on the heading title. For example, `# Hello World` would become `# Hello World {#hello-world}`.
   - To label an unnumbered heading, use `{.unnumbered}` at the end: 
     ```markdown
     # Hello World {-#nice-label}
     ```

2. **Reference the labeled heading**:  
   To reference a labeled heading anywhere in the text, use `\@ref(label)`. For example:
   ```markdown
   Please see Chapter \@ref(cross) for more details.
   ```
   - If you prefer text as the link instead of a numbered reference, you can use Markdown syntax:
     ```markdown
     [Click here to read about cross-references](#cross)
     ```

## Captioned Figures and Tables

Figures and tables with captions can also be cross-referenced from elsewhere in your book.

### Cross-Referencing Figures

To cross-reference a figure, add a chunk label and set the `fig.cap` option for the caption. Bookdown will automatically label the figure with `fig:chunk-label`.

Example:

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

You can reference this figure using `\@ref(fig:nice-fig)`. For example:

```markdown
See Figure \@ref(fig:nice-fig) for more details.
```

### Cross-Referencing Tables

To cross-reference a table, add a chunk label to the code chunk where you create the table with `knitr::kable()` and include a caption. Bookdown will automatically label the table with `tab:chunk-label`.

Example:

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can reference this table using `\@ref(tab:nice-tab)`. For example:

```markdown
Don't miss Table \@ref(tab:nice-tab).
```

## Summary

In this chapter, we covered how to create cross-references for headings, figures, and tables. Using cross-references helps readers navigate your document easily and allows you to create a more interactive and organized book.

Continue to the next chapter for more ways to enhance your Bookdown project.
